---
title: "Fit UGEE Models for Antibiotics Dataset"
author: "Ke Xu, MS<br><small>Department of Biostatistics<br>Vanderbilt University Medical Center</small>"
date: "<small>`r Sys.Date()`</small>"
knit: (function(inputFile, encoding) { 
      out_dir <- "/Users/jinyuanliu/Library/CloudStorage/OneDrive-VUMC/JL Method Paper/Big List/Aim3_Longitudinal_New/2025/SIM/Revision_July_2025/New Real Data/Share_with_JY_Oct_25";
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(out_dir, paste0('/Reports-UGEE-Complete-BC-',Sys.Date(),'.html'))) })
output: 
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: true
      smooth_scroll: false
    code_folding: hide
    theme: cerulean
css: rmarkdownstyle_x.css
---
<button id="toc-toggle" type="button" class="btn btn-default btn-xs code-folding-btn pull-right collapsed" onclick="toggletoc()"><span>Outline</span></button>
<script>
function toggletoc(){
$("#TOC").toggle();
}
</script>

***
**Overview**<br>
This report includes the results of UGEE model fitting using OTUs data at T1, and T2. Time -> Categorical, Group -> "MoCA low, Curiosity high" vs. 'Other'.


<!-- -->
**Variables**
MOCA: <=24 vs. >24. The missing values were replaced with random values within the range of this variable.
Curiosity: <=29 vs. >29. The missing values were replaced with random values within the range of this variable.
  
<!-- -->

***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
rm(list = ls())
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

library(readxl)
library(MGLM)
library(plyr)
library(vegan)
library(Rcpp)
library(MASS)
library(copula)
library(matrixStats)
library(dplyr)
library(biomformat)
library(readxl)
library(vegan)
library(dplyr)
library(stringr)
library(kableExtra)

specify_decimal <- function(x, k) trimws(format(round(x, k), nsmall=k))
```


# Load data
## Load OTU tables and meta data
```{r, warning=FALSE, message=FALSE}
# Import functions
sourceCpp("chisq_stat.cpp")
sourceCpp("dist2mat.cpp")

# UGEE
sourceCpp("ugee3_general_V2.cpp", showOutput = FALSE)

Data1 <- read.delim("genera.counts.tsv", 
                 header = TRUE,      # or FALSE if no header
                 stringsAsFactors = FALSE)
Meta1 <- read.delim("metadata.tsv", 
                 header = TRUE,      # or FALSE if no header
                 stringsAsFactors = FALSE)
Other1 <- read.delim("mtb.tsv", 
                 header = TRUE,      # or FALSE if no header
                 stringsAsFactors = FALSE)

otu_vars <- colnames(Data1)[-1]
```


```{r}
# recode timepoint and ID
master_dat <- full_join(Meta1, Data1)


master_dat$Timepoint <- str_extract(master_dat$Sample, "[^\\.]+$")
master_dat$ID <- str_extract(master_dat$Sample, "^[^\\.]+")
master_dat <- subset(master_dat, Timepoint != 'Flare')

```


## Subset data (Timepoint = 0, 1, 2, 3, 4)
```{r}
master_dat_sub <- master_dat[which(master_dat$Timepoint %in% c('0', '1', '2', '3', "4")), ] 

# select IDs with all timepoints
master_dat_sub <- master_dat %>%
  group_by(Subject) %>%                  
  filter(all(c(0,1,2,3,4) %in% Timepoint)) %>%  
  ungroup()


# t0
otu_t0 <- master_dat_sub %>%
  filter(Timepoint == '0') %>%
  dplyr::select(Sample, ID, Timepoint, Study.Group, any_of(otu_vars))

# t1
otu_t1 <- master_dat_sub %>%
  filter(Timepoint == '1') %>%
  dplyr::select(Sample, ID, Timepoint, Study.Group, any_of(otu_vars))

# t2
otu_t2 <- master_dat_sub %>%
  filter(Timepoint == '2') %>%
  dplyr::select(Sample, ID, Timepoint, Study.Group, any_of(otu_vars))

# t3
otu_t3 <- master_dat_sub %>%
  filter(Timepoint == '3') %>%
  dplyr::select(Sample, ID, Timepoint, Study.Group, any_of(otu_vars))

# t4
otu_t4 <- master_dat_sub %>%
  filter(Timepoint == '4') %>%
  dplyr::select(Sample, ID, Timepoint, Study.Group, any_of(otu_vars))

```


# Prepare parameters for function
## fij.mat.all
```{r}
#[ ] fij.mat.all.MAR
## T0
otu_t <- otu_t0[, -c(1:4)] # row = samples, col = otus
otu_t= otu_t/rowSums(otu_t) # covert to relative abundance
d_Y.t = vegdist(
  otu_t,
  method = "bray",
  binary = FALSE,
  diag = FALSE,
  upper = FALSE,
  na.rm = FALSE
) # beta diversity

fij.mat.all.t0.com = dist2mat(d_Y.t, 128) # transfer to a matrix


## T1
otu_t <- otu_t1[, -c(1:4)] # row = samples, col = otus
otu_t= otu_t/rowSums(otu_t) # covert to relative abundance
d_Y.t = vegdist(
  otu_t,
  method = "bray",
  binary = FALSE,
  diag = FALSE,
  upper = FALSE,
  na.rm = FALSE
) # beta diversity

fij.mat.all.t1.com = dist2mat(d_Y.t, 128) # transfer to a matrix

## T2
otu_t <- otu_t2[, -c(1:4)] # row = samples, col = otus
otu_t= otu_t/rowSums(otu_t) # covert to relative abundance
d_Y.t = vegdist(
  otu_t,
  method = "bray",
  binary = FALSE,
  diag = FALSE,
  upper = FALSE,
  na.rm = FALSE
) # beta diversity

fij.mat.all.t2.com = dist2mat(d_Y.t, 128) # transfer to a matrix

## T3
otu_t <- otu_t3[, -c(1:4)] # row = samples, col = otus
otu_t= otu_t/rowSums(otu_t) # covert to relative abundance
d_Y.t = vegdist(
  otu_t,
  method = "bray",
  binary = FALSE,
  diag = FALSE,
  upper = FALSE,
  na.rm = FALSE
) # beta diversity

fij.mat.all.t3.com = dist2mat(d_Y.t, 128) # transfer to a matrix

## T4
otu_t <- otu_t4[, -c(1:4)] # row = samples, col = otus
otu_t= otu_t/rowSums(otu_t) # covert to relative abundance
d_Y.t = vegdist(
  otu_t,
  method = "bray",
  binary = FALSE,
  diag = FALSE,
  upper = FALSE,
  na.rm = FALSE
) # beta diversity

fij.mat.all.t4.com = dist2mat(d_Y.t, 128) # transfer to a matrix


fij.mat.all <- list(fij.mat.all.t0.com,
                    fij.mat.all.t1.com,
                    fij.mat.all.t2.com,
                    fij.mat.all.t3.com,
                    fij.mat.all.t4.com)
```




## rij.all (no need)
```{r}
# no need, since we dont have any missing values

#[ ] rij.all
# 0 -> Missing; 1 -> completed

## T2
# Find rows with all missing values
# Ri_t2 = ori_dat.t2[ ,2]
# Ri_t2[!is.na(Ri_t2)] <- 1
# Ri_t2[is.na(Ri_t2)] <- 0
# rij.t2 = outer(Ri_t2, Ri_t2, FUN="*")


# rij.all <- list(rij.t2)
```


## grp_all (Group C, D, and H)

```{r}
set.seed(32608)
#[ ] grp_all
group.vars<-master_dat_sub$Study.Group[which(master_dat_sub$Timepoint == 0)]

group.num <- case_when(
  group.vars == "C" ~ 1,
  group.vars == "D" ~ 2,
  group.vars == "H" ~ 3
)
```

## z_mat1 (Age)
```{r}
#[ ] z_mat1
# list of between-subject attributes formed by continuous var # age
## T1
age.t1<-master_dat_sub$Age[which(master_dat_sub$Timepoint == 0)]

age = scale(age.t1)
```

## z_mat2 (BMI)
```{r}
#[ ] z_mat2
# list of between-subject attributes formed by continuous var # age
## T1
bmi.t1<-master_dat_sub$BMI[which(master_dat_sub$Timepoint == 0)]

bmi = scale(bmi.t1)

#sum(is.na(bmi.t1))
```

## Theta and Sigma_theta calculation
```{r}
fit.ugee = ugee3_general(dij_all = fij.mat.all, grp_all = group.num, 
                         z_ = as.matrix(cbind(age, bmi)),
                                  interaction_type = "modality") 
fit.ugee
```

# Define function for the Wald test
```{r}
library(MASS)  # for ginv

getwald <- function(results, terms) {
  betahat = results$theta[terms]
  sigma_betahat = results$Sigma_theta[terms, terms]
  
  Wstat = t(betahat) %*% ginv(sigma_betahat) %*% betahat
  df = length(terms)
  pval = pchisq(Wstat, df = df, lower.tail = FALSE)
  
  return(pval)
}
```

# test
```{r}
### Time Main Effect for Reference Group
getwald(fit.ugee, terms = c(2:5))

### Baseline Group Effect
getwald(fit.ugee, terms = c(6:10))

### Time * Group Interaction Effect
getwald(fit.ugee, terms = c(11:30))
```

# Slopes of CC
```{r}
base_CC = exp(fit.ugee$theta[1]) ## base mean
base_CC
base_CC*exp(fit.ugee$theta[c(2:5)]) - base_CC*c(1,exp(fit.ugee$theta[c(2:4)])) ## slopes
```

# Slopes of DD
```{r}
base_DD = exp(fit.ugee$theta[6]) * base_CC ## base mean
base_DD
base_CC*exp(fit.ugee$theta[c(11:14)]) - base_CC*exp(fit.ugee$theta[c(6,11:13)]) ## slopes
```

# Slopes of HH
```{r}
base_HH = exp(fit.ugee$theta[7]) * base_CC ## base mean
base_HH
base_CC*exp(fit.ugee$theta[c(15:18)]) - base_CC*exp(fit.ugee$theta[c(7,15:17)]) ## slopes
```

# location effect
```{r}
getwald(fit.ugee, terms = c(6:7)) ### Baseline location diff
getwald(fit.ugee, terms = c(11:14, 15:18)) ### location*time interaction
```

# scale effect
```{r}
getwald(fit.ugee, terms = c(8:10)) ### Baseline scale diff
getwald(fit.ugee, terms = c(19:30)) ### scale*time interaction
```

# scale&location interaction effect
```{r}
getwald(fit.ugee, terms = c(11:14, 15:18,19:30)) ### scale*time interaction
```

### Fit GLMM-MiraKIT
### prepare the data format
```{r}
##### prepare the data format ######
meta = data.frame(cbind(
      rep(group.num,5),
      rep(age,5),
      rep(bmi,5)))
n.group = length(unique(master_dat_sub$Subject))
meta$id = rep(c(1:n.group),5)
meta$visit = c(rep(1,n.group),rep(2,n.group),
               rep(3,n.group),rep(4,n.group),
               rep(5,n.group))
names(meta) = c("X_t","Z_1","Z_2","id","visit")
meta$X_t = meta$X_t-1
covs <- as.data.frame(cbind(meta$Z_1, meta$Z_2))
```

### Beta-div at 5 time points
```{r}
library(GLMMMiRKAT)
library(lme4)
library(Matrix)
library(permute)
### Beta-div at 5 time points
Matrix_OTU = rbind(otu_t0[, -c(1:4)],otu_t1[, -c(1:4)],
                   otu_t2[, -c(1:4)],otu_t3[, -c(1:4)],
                   otu_t4[, -c(1:4)])
if(requireNamespace("vegan")) {
  D.BC = as.matrix(vegdist(Matrix_OTU, method="bray"))
}
##### Create kernel matrices
if(requireNamespace("vegan")) {
  K.BC = D2K(D.BC)
}
mean(K.BC); sd(K.BC)
Ks = list(K.BC)

```

### Run GLMM-MiRKAT
```{r}
meta$X_t_binary_DD = ifelse(meta$X_t==1,1,0)
# Run GLMM-MiRKAT
GLMMMiRKAT = GLMMMiRKAT(meta$X_t_binary_DD, cov=covs, id=meta$id, Ks=Ks, model="binomial")
GLMMMiRKAT

otu_t0$Study.Group
```

### Run GLMM-MiRKAT
```{r}
meta$X_t_binary_CC = ifelse(meta$X_t==0,1,0)
# Run GLMM-MiRKAT
GLMMMiRKAT = GLMMMiRKAT(meta$X_t_binary_CC, cov=covs, id=meta$id, Ks=Ks, model="binomial")
GLMMMiRKAT
```

### Git UGEE
```{r}
group_binary_DD = ifelse(group.num == 2, 1, 0)
fit.ugee_DD = ugee3_general(dij_all = fij.mat.all, grp_all = group_binary_DD, 
                         z_ = as.matrix(cbind(age, bmi)),
                                  interaction_type = "modality") 
fit.ugee_DD

# scale&location interaction effect
getwald(fit.ugee_DD, terms = c(2:17)) ### scale&location omnibus
```

### Git UGEE
```{r}
group_binary_CC = ifelse(group.num == 1, 1, 0)
fit.ugee_CC = ugee3_general(dij_all = fij.mat.all, grp_all = group_binary_CC, 
                         z_ = as.matrix(cbind(age, bmi)),
                                  interaction_type = "modality") 
fit.ugee_CC

# scale&location interaction effect
getwald(fit.ugee_CC, terms = c(2:17)) ### scale&location omnibus
```